---
title: "02-simulation-for-GLM(M)s"
format: 
  html:
    math: mathjax
editor: 
  markdown: 
    wrap: 100
---

```{r}
# load relevant libraries
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
```

## Simulation for GLM(M)s

How many times have you sat with your data, googling, asking ChatGPT, pulling hair out trying to
figure out which model to fit for your experiment/study? If you're like me, it's been a lot of
times. How many times have you sat there thinking: Does this model even do what I think it's doing?
Again, if you're like me, there have been many times.

Now, what if I told you there was a way to solve these kinds of problems, improve reproducibility
and your statistics skills? That solution is simulation.

### Example N: Biodiversity in food forests

There was a lively debate in our department recently about the correct way to analyse an experiment
that examined how the *food-forest* land-use type affects biodiversity. Specifically, the
researchers were interested in whether food forests had more biodiversity value than other land-use
types (e.g. agricultural fields). The experiment was designed in the following way. There were 30
locations throughout the Netherlands and Flanders (Belgium). At each of these 30 locations, there
were two plots from the two groups i.e. food forest and agricultural field (there were actually four
groups but, for simplicity, we will use two for this example). Within each of plots at each of the
30 locations (i.e. $30 * 2 = 60$ plots), ten samples were randomly taken. Therefore, within each of
the two plots at a given location, 10 individual biodiversity measurements were taken (i.e.
$30 * 2 * 10 = 600$ measurements). For this analysis, we will view the agricultural field group as
the *control* and the food forest group as the *treatment* but this could also reasonably be
switched around. In this specific case, it does not matter.

**Exercise**: How would you analyse these data?

The resarcher proposed analysing these data using a linear mixed model (for simplicity, we will use
a Gaussian error distribution but the same ideas apply to generalised linear models) with the
following structure: biodiversity (B) \~ treatment (T) + (1 \| location (L) (I assume many of you
have seen something like this before).

But is this correct way to analyse these data? Let's think very carefully about what this model is
assuming. Before we do this, we will write the model equation. Do not worry about this model
equation for now. However, I wrote it here because we will want to come back to all the different
parameters in this model.

$$
B_{ij} \sim Normal(\mu_{ij}, \sigma_{residual})\\
\mu_{ij} = a_{j} + \beta_1(treatment_{i})\\
\alpha_{j} \sim Normal(\mu_{global}, \sigma_{global}) \text{, for location j = 1,} \dots \text{,J}
$$

So, what does this model assume? It assumes that the control plots (i.e. the agricultural fields) at
all 30 locations ($j$) have some base-level of biodiversity ($\alpha_j$). These base-level
biodiversity values ($\alpha_j$) are normally distributed with some global mean ($\mu_{global}$) and
global standard deviation ($\sigma_{global}$). The model then assumes that the treatment has the
same effect in all 30 locations (i.e. the difference between food forest and agricultural fields is
the same across the 30 locations). Based on the base-level biodiversity values at the 30 locations
($\alpha_j$) and the treatment effect ($\beta_{1}$), the control (i.e. agricultural field) and
treatment group (i.e. food forest) at each location have some mean value of biodiversity
($\mu_{ij}$). The observed biodiversity values ($B_{ij}$) are normally distributed around these mean
values ($\mu_{ij}$) for each location-group combination with some residual standard deviation
($\sigma_{residual}$).

Now we are going to simulate from this model and, I promise, this will become clearer. We went
painfully through all the parameters in the model because the idea behind using simulation is to
simulate these assumptions with specific parameter values, fit the model and then see if our model
reproduces these parameter values. If the model cannot reproduce the parameters, then it means that
our assumptions about how the model works are probably wrong.

#### Simulation 1

To start, we will specify the global mean ($\mu_{global}$) and the global standard deviation
($\sigma_{global}$) of the normal distribution that describes the base-level of biodiversity at each
location in the control group (i.e. the agricultural field). Let's visualise this:

```{r}
# set-up the global mean (mu_global = 100) and global standard deviation (sigma_global = 10)
mu_global <- 100
sigma_global <- 10

# sample from this distribution and plot the results
dist_global <- rnorm(n = 100000, mean = mu_global, sd = sigma_global)

ggplot(mapping = aes(x = dist_global)) +
  geom_density(fill = "grey", bw = 2.5) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  ylab("Density") +
  xlab("Biodiversity") +
  theme_minimal()
```

In this experiment, we have 30 different locations. Therefore, we need to sample 30 times from this
global distribution. These 30 samples are then the 30 base-level biodiversity values for the
agricultural field plot (i.e. control plot) at each location ($\alpha_j$). We then visualise these
30 location mean values as red vertical lines on the global distribution:

```{r}
# set the number of locations
locations <- 30

# take 15 random samples from the distribution
a_j <- rnorm(n = locations, mean = mu_global, sd = sigma_global)

ggplot() +
  geom_density(mapping = aes(x = dist_global),
               fill = "grey", bw = 2.5) +
  geom_vline(mapping = aes(xintercept = a_j), colour = "red", size = 0.25) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  ylab("Density") +
  xlab("Biodiversity") +
  theme_classic()
```

So, now we have our 30 base-level biodiversity values for agricultural fields at the 30 locations
($\alpha_j$). Let's visualise this in a different way that is more intuitive. Specifically, we plot
each mean value on the y-axis and the x-axis simply indexes the 30 locations. Moreover, we plot the
global mean as a red, dashed, horizontal line.

```{r}
# visualise the results as a location scatterplot
ggplot() +
  geom_point(mapping = aes(x = seq_along(a_j), y = a_j),
             colour = "black", size = 2) +
  geom_hline(yintercept = mu_global, colour = "red", linetype = "dashed") +
  ylab("alpha j") +
  xlab("Location j") +
  theme_classic()
```

The next thing that this model assumes is that the base-level biodiversity value in the control
group (i.e. the agricultural field) is modified by the treatment (i.e. the food forest treatment
group) ($\beta_1$. Moreover, in this specific model formulation, the treatment effect is assumed to
be constant across all 30 locations.

We'll assume a treatment effect of 5 (i.e. $\beta_1 = 5$). This means that, on average, biodiversity
in food forests will be greater than in the agricultural fields by 5 units (corresponds to 5%). We
then add this treatment effect to all locations which gives us the mean biodiversity value at the 30
locations for the food forest.

```{r}
# set the treatment effect
b1 <- 5

# add the treatment effect
mu_ij <- a_j + b1

# add the control plots to this vector
mu_ij <- c(a_j, mu_ij)

# pull into a data.frame
dat_mu <- dplyr::tibble(location = rep(c(seq_len(locations)), 2),
                        treatment = rep(c("agricultural: control", "food-forest: treatment"), each = locations),
                        mu_ij = mu_ij)

# visualise the results as a location scatterplot
ggplot(data = dat_mu,
       mapping = aes(x = location, y = mu_ij, colour = treatment)) +
  geom_point() +
  geom_hline(yintercept = mu_global, colour = "red", linetype = "dashed") +
  ylab("mu ij") +
  xlab("Location j") +
  theme_classic()
```

So, now we have simulated the average biodiversity values in the control group (i.e. agricultural
fields) and the average biodiversity values in the treatment group (i.e. food forests) as per the
model (we hope). The last thing we need to do is simulate the *observed biodiversity value* i.e. the
10 replicates in each plot in each location ($B_i$).

To do this, we assume that the biodiversity values ($B_i$) are normally distributed around the means
($\mu_{ij}$) with some standard deviation ($\sigma_{residual}$) which we will set at 5:

```{r}
# number of within-plot replicates
replicates <- 10

# set the residual standard deviation
sigma_residual <- 5

# simulate six replicates in each location-group combination
dat_list <- vector("list", length = nrow(dat_mu)) 
for (i in seq_along(dat_list)) {
  
  # extract the first location-group combination
  x <- dat_mu[i, ]
  
  # simulate 6 random draws using the mean and residual standard deviation
  y <- rnorm(n = replicates, mean = x$mu_ij, sd = sigma_residual)
  
  # add this to the data.frame
  z <- dplyr::tibble(location = x$location,
                     treatment = x$treatment,
                     B_i = y)
  
  # add this to an output list
  dat_list[[i]] <- z
  
}

# bind into a data.frame
dat <- dplyr::bind_rows(dat_list)

```

Now we have a fully simulated dataset which we can plot and which shows the simulated mean values as
diamonds ($\mu_{ij}$) and the simulated observed biodiversity values as circles ($B_i$).

```{r}
# visualise the results as a location scatterplot
ggplot() +
  geom_point(data = dat_mu,
             mapping = aes(x = location, y = mu_ij, colour = treatment), 
             size = 3.5, shape = 18) +
  geom_point(data = dat,
             mapping = aes(x = location, y = B_i, colour = treatment),
             alpha = 0.1) +
  geom_hline(yintercept = mu_global, colour = "red", linetype = "dashed") +
  ylab("mu ij") +
  xlab("Location j") +
  theme_classic()
```

Using this simulated dataset, we can fit the linear mixed model using the lme4 package and see if we
can reproduce the parameter values that we put into the model.

```{r}
# check the structure of the data
str(dat)

# convert the location variable to a character
dat$location <- as.character(dat$location)

# fit the model
lmm1 <- lmerTest::lmer(B_i ~ treatment + (1 | location), data = dat)

# extract the summary
lmm1_sum <- summary(lmm1)

# print the summary
print(lmm1_sum)
```

Now that we have fit the model, we can see if we can reproduce the parameters that we put into the
model. Let's start with the global mean ($\mu_{global}$) and global standard deviation
($\sigma_{global}$).

We set the global mean ($\mu_{global}$) to 100. Therefore, we should obtain a value close to 100
from our fitted model. To extract the mean, we need the intercept coefficient in this model. In this
model, the estimated global mean was indeed very close to 100 (see below):

```{r}
# extract the global mean from the model: intercept term
print(paste0("Estimated global mean: ", lmm1_sum$coefficients[1,1]))
```

What about the global standard deviation ($\sigma_{global}$) which we set at 10? Indeed, the
estimated global standard deviation is 9.4, very close to the 10 that we simulated:

```{r}
# extract the random effects
lmm1_re <- VarCorr(lmm1)

# the relevant standard deviation is the location standard deviation
x <- attr(lmm1_re$location, "stddev")
names(x) <- NULL

# extract the global mean from the model: intercept term
print(paste0("Estimated global standard deviation: ", x))
```

What about the treatment effect ($B_1$)? We simulated a treatment effect of 5. Does the model
reproduce this? Yes, it does.

```{r}
# extract the treatment effect from the model
print(paste0("Estimated treatment effect: ", lmm1_sum$coefficients[2,1]))
```

And, what about the residual variation which we set at 5? This also matches very well with our
simulation:

```{r}
# extract the global mean from the model: intercept term
print(paste0("Estimated global standard deviation: ", lmm1_sum$sigma))
```

Finally, the model estimates a mean value for each location ($\alpha_j$). We can also check if these
mean values match with simulated the mean values for each site.

```{r}
# extract the model coefficients
lmm1_coef <- coef(lmm1)
lmm1_coef <- lmm1_coef$location

# get the rowname as a nummeric variable and sort
lmm1_coef$location <- as.numeric(row.names(lmm1_coef))
lmm1_coef <- dplyr::arrange(lmm1_coef, location)

# how do these values correlate with the simulated values
plot(a_j, lmm1_coef$`(Intercept)`)

```

Hopefully, this exercise has convinced you that we understand the behaviour of the model because we
were able to directly simulate it and the model reproduced all the simulated parameters with high
levels of accuracy. Also, it is worth going back and increasing the replication in this simulation
using the *locations* and *replicates* objects. If you increase these, the estimates should get
better and better.

But, now that we understand the assumptions and behaviour, we may ask ourselves are these
assumptions reasonable?

**Exercise: Are there any assumptions about how this model works that you do not think make much
sense?**

#### Simulation 2

In my view, the assumption that the treatment effect is the same across the different sites is
problematic for this kind of study. No food forest is the same and no agricultural field is the
same. Moreover, the locations cover a huge area with different environmental conditions which means
that the biodiversity difference between a food forest and a grassland might be very different. By
only fitting a single treatment effect parameter, the model is ignoring all of that variation and
this can cause the significance of the effect to be inflated (Barr et al. 2013; Gelman and Brown
2024). Rather, what we need to do, is model the potential variation in treatment effects among the
locations. By doing this, we will obtain the average treatment effect across all 15 locations by
directly modelling variation among locations.

So what is different about this model compared to the previous model? Well, the only thing that
differs is that we are now fitting a separate intercept and treatment effect for each location
rather than simply a separate intercept for each location and a single treatment effect. Moreover,
we will assume that the 15 intercepts and treatment effects are correlated. This just means that,
for example, the model allows for the possibility that, for example, a high intercept is typically
associated with a high (or low) treatment effect. As previously, I'm going to write the model
equation out and then we will explore the assumptions of this model when we perform the simulation.

$$
B_{ij} \sim Normal(\mu_{ij}, \sigma_{\text{residual}})\\
\mu_{ij} = \alpha_{j} + \beta_{j}(\text{treatment}_{i})\\
\begin{pmatrix}
\alpha_{j} \\
\beta_{j}
\end{pmatrix}
\sim \mathcal{MVN} \left(
\begin{pmatrix}
\bar{\alpha} \\
\bar{\beta}
\end{pmatrix},
\begin{pmatrix}
\tau_{\alpha}^2 & \rho \tau_\alpha \tau_\beta \\
\rho \tau_\alpha \tau_\beta & \tau_{\beta}^2
\end{pmatrix}
\right), \text{for location } j = 1, \dots, J
$$

So, what does this model assume? It assumes that the control plots (i.e. the agricultural fields) at
all 15 locations ($j$) have some base-level of biodiversity ($\alpha_j$). This the same as the
previous model. The model then assumes that the treatment (i.e. the difference between the food
forest and the agricultural fields) differs across the 15 locations such that there is a separate
treatment effect at each location ($\beta_{j}$). Therefore, based on the base-level biodiversity
values at the 15 locations ($\alpha_j$) and the treatment effects at the 15 locations ($\beta_{j}$),
the control (i.e. agricultural field) and treatment group (i.e. food forest) at each location have
some mean value of biodiversity ($\mu_{ij}$). Importantly, the $\alpha_j$ and $\beta_j$ values come
from a common multivariate normal distribution. The observed biodiversity values ($B_{ij}$) are
normally distributed around these mean values ($\mu_{ij}$) for each location-group combination with
some residual standard deviation ($\sigma_{residual}$).

To start with, we need to define the parameters of the multivariate normal distribution. As with the
previous simulation, we will assume that the global mean across locations is 100 ($\bar{\alpha_j}$)
and that the standard deviation is 10 (variance is 100), ($\tau_{\alpha}^2$). Next, we assume that
average treatment effect is 5 ($\bar{\beta_j}$) and that the standard deviation is 2.5 (variance =
6.25) ($\tau_{\beta}^2$). Next, we assume that the $\alpha_j$ and $\beta_j$ values are correlated
across locations with a correlation coefficient of 0.3 ($\rho$). Using these simulated values, we
can draw pairs of $\alpha_j$ and $\beta_j$ values for each location from a multivariate normal
distribution.

```{r}
# sim global mean (alpha_bar = 100) and variance (tau_alpha = 10)
alpha_bar <- 100
tau_alpha <- 10

# sim global treatment effect mean (beta_bar = 5) and standard deviation (tau_beta = 2.5)
beta_bar <- 5
tau_beta <- 3

# set the correlation coefficient
rho <- 0.5

# build the covariance matrix
cov_mat <- matrix(data = c((tau_alpha^2), 
                           rho*tau_alpha*tau_beta, 
                           rho*tau_alpha*tau_beta, (tau_beta^2)), 2, 2)

# simulate from the multivariate normal
dist_global <- MASS::mvrnorm(n = 20000,
                             mu = c(alpha_bar, beta_bar),
                             Sigma = cov_mat)

# plot this multivariate normal distribution
dist_global <- dplyr::as_tibble(as.data.frame(dist_global))
names(dist_global) <- c("a_j", "b_j")

ggplot(data = dist_global,
       mapping = aes(x = a_j, y = b_j)) +
  geom_point() +
  ylab("beta_j") +
  xlab("alpha_j") +
  theme_minimal()
```

This figure shows 20000 samples from the multivariate normal distribution that we set-up for
$\alpha_j$ and $\beta_j$. As you can see, there values are correlated such that information from
$\alpha_j$ can provide information about $\beta_j$. As this is the general distribution, we need to
sample 15 locations from this which we overlay as red dots where each point represents both an
$\alpha_j$ and $\beta_j$ for each location.

```{r}
# set the number of locations
locations <- 30

# simulate from the multivariate normal
dist_samp <- MASS::mvrnorm(n = locations,
                           mu = c(alpha_bar, beta_bar),
                           Sigma = cov_mat)

# plot this multivariate normal distribution
dist_samp <- dplyr::as_tibble(as.data.frame(dist_samp))
names(dist_samp) <- c("a_j", "b_j")

ggplot() +
  geom_point(data = dist_global,
       mapping = aes(x = a_j, y = b_j)) +
  geom_point(data = dist_samp,
             mapping = aes(x = a_j, y = b_j), colour = "red", size = 3.5) +
  ylab("beta_j") +
  xlab("alpha_j") +
  theme_minimal()
```

Using these sampled values of $\alpha_j$ and $\beta_j$, we can now build a dataset with the means of
each control and treatment plot at each location ($\mu_{ij}$). Let's visualise this.

```{r}
# build a data.frame with the simulated means
dat_mu <- 
  dplyr::tibble(location = rep(c(seq_len(locations)), 2),
                treatment = rep(c("agricultural: control", "food-forest: treatment"),
                                each = locations),
                mu_ij = c(dist_samp$a_j, (dist_samp$a_j + dist_samp$b_j)))
head(dat_mu)

# visualise the results as a location scatterplot
ggplot(data = dat_mu,
       mapping = aes(x = location, y = mu_ij, colour = treatment)) +
  geom_point() +
  ylab("mu ij") +
  xlab("Location j") +
  theme_classic()
```

As with the previous model, now that we have the simulated the average biodiversity values in the
control group (i.e. agricultural fields) and the average biodiversity values in the treatment group
(i.e. food forests), the last thing we need to do is simulate the *observed biodiversity value* i.e.
the 6 replicates in each plot in each location ($B_i$).

To do this, we assume that the biodiversity values ($B_i$) are normally distributed around the means
($\mu_{ij}$) with some standard deviation ($\sigma_{residual}$) which we will set at 5:

```{r}
# number of within-plot replicates
replicates <- 10

# set the residual standard deviation
sigma_residual <- 5

# simulate six replicates in each location-group combination
dat_list <- vector("list", length = nrow(dat_mu)) 
for (i in seq_along(dat_list)) {
  
  # extract the first location-group combination
  x <- dat_mu[i, ]
  
  # simulate 6 random draws using the mean and residual standard deviation
  y <- rnorm(n = replicates, mean = x$mu_ij, sd = sigma_residual)
  
  # add this to the data.frame
  z <- dplyr::tibble(location = x$location,
                     treatment = x$treatment,
                     B_i = y)
  
  # add this to an output list
  dat_list[[i]] <- z
  
}

# bind into a data.frame
dat <- dplyr::bind_rows(dat_list)

```

```{r}
# fit the model with correlated random intercepts and slopes
lmm2 <- lmerTest::lmer(B_i ~ treatment + (1 + treatment | location), data = dat)
summary(lmm2)
```

### Functions to repeat these simulations

Now, we want to test the behaviour of these models over many replicate simulations. To do this, I
have wrapped the code for simulation 2 into a function that will allow us to do this. The code is
very similar but is now more streamlined and allows one to easily run the simulation with different
parameter values.

```{r}
# sim_lmm: fit lmm1 and lmm2 to simulated data
sim_lmm <- function(alpha_bar = 100, tau_alpha = 10, beta_bar = 5, tau_beta = 3, 
                     rho = 0.5, locations = 30, replicates = 10, sigma_residual = 5) {
  
  # Load required libraries
  if (!requireNamespace("dplyr", quietly = TRUE)) stop("Package 'dplyr' is required.")
  if (!requireNamespace("lmerTest", quietly = TRUE)) stop("Package 'lmerTest' is required.")
  if (!requireNamespace("MASS", quietly = TRUE)) stop("Package 'MASS' is required.")
  
  # Build the covariance matrix
  cov_mat <- matrix(c(tau_alpha^2, rho * tau_alpha * tau_beta, 
                      rho * tau_alpha * tau_beta, tau_beta^2), 
                    nrow = 2, ncol = 2)
  
  # Simulate from the multivariate normal distribution
  dist_samp <- MASS::mvrnorm(n = locations, mu = c(alpha_bar, beta_bar), Sigma = cov_mat)
  
  # Convert to tibble and name columns
  dist_samp <- dplyr::as_tibble(as.data.frame(dist_samp))
  names(dist_samp) <- c("a_j", "b_j")
  
  # Build the data frame with simulated means
  dat_mu <- 
    dplyr::tibble(
    location = rep(seq_len(locations), 2),
    treatment = rep(c("control", "treatment"), each = locations),
    mu_ij = c(dist_samp$a_j, dist_samp$a_j + dist_samp$b_j)
  )
  
  # Simulate replicates for each location-treatment combination
  dat <- 
    dat_mu |> 
    dplyr::rowwise() |>
    dplyr::mutate(
      B_i = list(rnorm(n = replicates, mean = mu_ij, sd = sigma_residual))
    ) |> 
    dplyr::select(location, treatment, B_i) |>
    tidyr::unnest(cols = c(B_i))
  
  # Convert location variable to character
  dat$location <- as.character(dat$location)
  
  # Fit the linear mixed model with a random intercept only
  lmm1 <- lmerTest::lmer(B_i ~ treatment + (1 | location), data = dat)
  
  # Fit the linear mixed model with correlated random intercepts and slopes
  lmm2 <- lmerTest::lmer(B_i ~ treatment + (1 + treatment | location), data = dat)
  
  # Return the simulated data and the fitted model
  list(model1 = lmm1,
       model2 = lmm2)
  
}
```

So, let's see how this function works. We will run a single replicate simulation (i.e. simulation 2
where treatment effects vary across locations) and fit both models (model1 and model2 respectively).
This should convince you that the models are similar to what we have already done. Moreover, feel
free to adjust some of the parameters and see what happens. For example, if you increase the number
of locations (i.e. *locations*) and the number of replicates within each location (i.e.
*replicates*), you should get estimates that are closer to the true, simulated parameter values.

```{r}
# run one replicate of the second simulation
sim2_obj <- sim_lmm(alpha_bar = 100, tau_alpha = 10, beta_bar = 5, tau_beta = 3, 
                    rho = 0.3, locations = 30, replicates = 10, sigma_residual = 5)
summary(sim2_obj$model1)
summary(sim2_obj$model2)
```

### Comparison of the two models

What was our question at the start? Well, we wanted to determine which model we should use (model 1
i.e. random intercept or model 2 i.e. correlated random intercept and random slope). I hope I've
convinced you that the assumptions of model 1 that the treatment effect is the same across all
locations is not very realistic and that model 2, which allows the treatment effect to vary across
locations is more realistic. Now, let's use simulation 2 where the treatment effect varies across
locations and compare the inference we get from the two models. To do this, we will simulate 1000
replicates of simulation 2 and compare the resulting t-values for the treatment effect for models 1
and 2.

```{r}
sim_reps <- vector("list", length = 1000)
for (i in seq_along(sim_reps)) {

  # run one replicate of the second simulation (i.e. correlated random intercepts and slopes)
  sim_obj <- sim_lmm(alpha_bar = 100, tau_alpha = 10, beta_bar = 5, tau_beta = 3, 
                     rho = 0.3, locations = 30, replicates = 10, sigma_residual = 5)
  
  # extract the model objects
  sim1_sum <- summary(sim_obj$model1)
  sim2_sum <- summary(sim_obj$model2)
  
  # extract the t-values for the treatment
  sim1_t <- coefficients(sim1_sum)[2, 4]
  sim2_t <- coefficients(sim2_sum)[2, 4]
  
  # extract the p-value for the treatment
  sim1_p <- coefficients(sim1_sum)[2, 5]
  sim2_p <- coefficients(sim2_sum)[2, 5]
  
  # bind into a data.frame
  sim_reps[[i]] <- dplyr::tibble(rep = i,
                                 sim = c(1, 2),
                                 t_val = c(sim1_t, sim2_t),
                                 p_val = c(sim1_p, sim2_p))
  
}

# bind into a data.frame
sim_rep_df <- dplyr::bind_rows(sim_reps)
```

Now, we can compare the t-values between the two models on the same simulated data where the
treatment effect varied between locations. What we see is that the t-values are consistently higher
in model 1. This is because the model is ignoring the variation among locations. Model 2 is less
certain about the treatment effect because it *knows* that there is variation. Indeed, the fact that
the t-value for the treatment effect in model 1 is extremely high (on average 12) should raise some
alarm bells.

```{r}
# compare the t-values between the two simulations
ggplot(data = sim_rep_df,
       mapping = aes(x = t_val, fill = as.character(sim))) +
  geom_density(alpha = 0.5, bw = 0.7) +
  theme_classic()
```

In my view, the only time when model 1 would be appropriate is if there is only one replicate per
site in which case the experiment becomes a randomised block design. The model would then not be
able to properly estimate the varying slope effects across locations. The other case is if there is
no variation among locations in the treatment effect. Let's do this by simply setting the slope
variation to zero. What happens in this case, is that the more complex model 2 becomes very similar
to model 1. Moreover, the model will probably complain because of some numerical problems (e.g.
boundary fit singular). This occurs because the model is trying to estimate the variation in beta
but it is simply not there.

```{r}
sim_reps <- vector("list", length = 1000)
for (i in seq_along(sim_reps)) {

  # run one replicate of the second simulation (i.e. correlated random intercepts and slopes)
  sim_obj <- sim_lmm(alpha_bar = 100, tau_alpha = 10, beta_bar = 5, tau_beta = 0, 
                     rho = 0, locations = 30, replicates = 10, sigma_residual = 5)
  
  # extract the model objects
  sim1_sum <- summary(sim_obj$model1)
  sim2_sum <- summary(sim_obj$model2)
  
  # extract the t-values for the treatment
  sim1_t <- coefficients(sim1_sum)[2, 4]
  sim2_t <- coefficients(sim2_sum)[2, 4]
  
  # extract the p-value for the treatment
  sim1_p <- coefficients(sim1_sum)[2, 5]
  sim2_p <- coefficients(sim2_sum)[2, 5]
  
  # bind into a data.frame
  sim_reps[[i]] <- dplyr::tibble(rep = i,
                                 sim = c(1, 2),
                                 t_val = c(sim1_t, sim2_t),
                                 p_val = c(sim1_p, sim2_p))
  
}

# bind into a data.frame
sim_rep_df <- dplyr::bind_rows(sim_reps)
```

```{r}
# compare the t-values between the two simulations
ggplot(data = sim_rep_df,
       mapping = aes(x = t_val, fill = as.character(sim))) +
  geom_density(alpha = 0.5, bw = 0.7) +
  theme_classic()
```

All of this is to say that, in this case, these simulations indicate that using model 2 is a more
appropriate analysis. The only time that model 1 would become more appropriate is if there was only
a single replicate per site or if there was very little variation in the treatment effect among
locations. This is almost certainly not the case given this kind of field experiment. Moreover, even
if this did turn out to be the case, it would still be better to try and model the variation in
treatment effects and, only if the model was unable to fit, to use the simpler model.

This post was inspired by two great papers that helped me understand this varying intercept/varying
slope distinction which are both well-worth reading. They are talking from a psychology perspective
but similar problems apply to the ecological literature:

-   Gelman and Brown (2024): http://www.stat.columbia.edu/\~gelman/research/published/healing3.pdf
-   Barr et al. (2013): https://www.sciencedirect.com/science/article/pii/S0749596X12001180

### Bonus round...

In this case, I made the simplifying assumption that the biodiversity variable was continuous and
normally distributed. However, very few measures of biodiversity are actually like this... Take
species richness, for example, this is probably more appropriate modelled as a Poisson random
variable. We are now going to repeat Simulation 2 but using a Poisson generalised linear mixed model
to show you what this looks like.

Like previously, we are going to write the equation. As you can, the model is very similar. The only
difference is that the observed biodiversity ($B_{ij}$) is Poisson distributed around some
($\lambda_{ij}$). Moreover, the $\alpha_{j}$ and $\beta_{j}$ parameters determine the
$log(\lambda_{ij}$ value. This ensures that the $\lambda_{ij}$ values are positive and that the
linear model goes from negative infinity to positive infinity.

$$
B_{ij} \sim \text{Poisson}(\lambda_{ij}) \\
\log(\lambda_{ij}) = \alpha_{j} + \beta_{j}(treatment_{i}) \\
\begin{pmatrix}
\alpha_{j} \\
\beta_{j}
\end{pmatrix}
\sim \mathcal{MVN} \left(
\begin{pmatrix}
\bar{\alpha} \\
\bar{\beta}
\end{pmatrix}
,
\begin{pmatrix}
\tau_{\alpha}^2 & \rho \tau_\alpha \tau_\beta \\
\rho \tau_\alpha \tau_\beta & \tau_{\beta}^2
\end{pmatrix}
\right), \text{ for location } j = 1, \dots, J
$$

Given the similarity in the equation, the simulation is very similar as well.

```{r}
# set the mean species richness value across locations
alpha_bar = log(100)

# set the variation in species richness across locations
tau_alpha = log(100)/10
print(tau_alpha)

# set the treatment effect
beta_bar = log(5)

# set the variation in the treatment effect
tau_beta = log(5)/2

# set the correlation between beta_j and alpha_j
rho = 0.5

# set the number of locations
locations = 1000

# set the number of replicates within each site (two sites per location)
replicates = 50
  
# Build the covariance matrix
cov_mat <- matrix(c(tau_alpha^2, rho * tau_alpha * tau_beta, 
                  rho * tau_alpha * tau_beta, tau_beta^2), 
                  nrow = 2, ncol = 2)
  
# Simulate from the multivariate normal distribution
dist_samp <- MASS::mvrnorm(n = locations, mu = c(alpha_bar, beta_bar), Sigma = cov_mat)
  
# Convert to tibble and name columns
dist_samp <- dplyr::as_tibble(as.data.frame(dist_samp))
names(dist_samp) <- c("a_j", "b_j")
  
# Build the data frame with simulated means
dat_mu <- 
  dplyr::tibble(
  location = rep(seq_len(locations), 2),
  treatment = rep(c("control", "treatment"), each = locations),
  mu_ij = c(dist_samp$a_j, dist_samp$a_j + dist_samp$b_j)
  )

```

But, things change when we have to simulated the observed biodiversity values. For this, we take the
exponentiated $\mu_{ij}$ value and pass it through the Poisson distribution to generate the observed
biodiversity values ($B_{ij}$).

```{r}
# Simulate species richness replicates for each location-treatment combination
dat <- 
  dat_mu |> 
  dplyr::rowwise() |>
  dplyr::mutate(
    B_i = list(rpois(n = replicates, lambda = exp(mu_ij)))
    ) |> 
  dplyr::select(location, treatment, B_i) |>
  tidyr::unnest(cols = c(B_i))
  
# Convert location variable to character
dat$location <- as.character(dat$location)
  
# Fit the linear mixed model with correlated random intercepts and slopes
glmm1 <- lme4::glmer(B_i ~ treatment + (1 + treatment | location), 
                     family=poisson(link = log), data = dat)

# check the object
summary(glmm1)
```
